---
title: "Generating a Vocabulary and Lookup"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generating a Vocabulary and Lookup}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  eval = FALSE,
  comment = "#>"
)
```

```{r setup}
# Install non-CRAN packages
# remotes::install_github("macmillancontentscience/wikimorphemes")
# remotes::install_github("macmillancontentscience/wordpiece.data")
library(morphemepiece)
library(wikimorphemes)
library(wordpiece.data)
library(dplyr)
library(ggplot2)
library(purrr)
```

This vignette shows how to use the processed words from the {wikimorphemes}
package to create a morphemepiece vocabulary and lookup table.

To make a morphemepiece vocabulary, start with...

 - a wordpiece vocabulary
 - a processed wikimorphemes cache
 - (optionally) a word frequency table derived from some corpus.

```{r load-data}
# load functions to make the vocab + lookup
source(here::here("vignettes", "make_vocab_and_lookup.R"))

# Load data from various packages related to this task.
original_lookup <- readRDS(wikimorphemes::download_wikimorphemes_lookup())

# TODO: Add something to reproduce this or host it somewhere.
word_frequency_table <- readRDS(
  fs::path(
    morphemepiece_cache_dir(),
    "word_frequency_table.rds"
  )
)

# Not all wiktionary words are in the wiktionary lookup. Use the full word list
# to add short words back in.
full_lookup <- .add_words_to_lookup(
  original_lookup, 
  wikimorphemes::wiktionary_word_list()
)
```

Before we make a specific vocabulary, we can look at which morphemes are most
common over all wiktionary words to get a prioritized list.

```{r count-tokens}
# for much of this process, it's more convenient to have the processed words
# unnested, with one morpheme per row. This takes a few minutes.

# Currently, we're considering only words with pure lowercase latin characters.
# We likely will want to also include the simplified version of words with
# accented characters in this list. 
unnested_lookup <- .unnest_lookup(full_lookup, clean = TRUE)

# count how many wiktionary words each token appears in.
token_counts <- count_tokens(unnested_lookup)

utils::head(token_counts)
```

Unsurprisingly, the top token is "##s", the "s" inflection at the ends of words.

We rank tokens in order of frequency. 
We can also rank *words* in order of the maximum rank that their component tokens have. 
This shows which words would be covered by a token vocabulary including tokens up to some rank.

```{r temp-values, eval = TRUE, include = FALSE}
# This is here so the package can still build properly when this vignette is
# "turned off."
words_vs_tokens <- list(
  max_rank = 194347L,
  n_words = 643745L
)
```

```{r token-ranks}
# Some words process into "non-clean" tokens (diacrits, etc.).
# Those tokens are excluded in token_counts, so will get an NA here.
# Deal with this better later, but for now, just remove those words.
# (only about 0.01% of words)

# Find the highest-rank (rarest) token within each word.
words_with_max_token_ranks <- dplyr::left_join(
  unnested_lookup, 
  token_counts, 
  by = "token"
) %>% 
  dplyr::group_by(word) %>% 
  dplyr::summarize(max_rank = max(rank)) %>% 
  dplyr::filter(!is.na(max_rank))

# Count how many total words are covered by tokens up to some rank:
words_vs_tokens <- words_with_max_token_ranks %>% 
  dplyr::group_by(max_rank) %>% 
  dplyr::summarize(n_words = dplyr::n_distinct(word)) %>% 
  dplyr::arrange(max_rank) %>% 
  dplyr::mutate(n_words = cumsum(n_words))
```

There are `r max(words_vs_tokens$max_rank)` distinct tokens and `r max(words_vs_tokens$n_words)` distinct words.

```{r plot-counts}
# plot!
words_vs_tokens %>% 
  dplyr::mutate(frac_words = n_words/max(n_words)) %>% 
  dplyr::mutate(frac_tokens = max_rank/max(max_rank)) %>% 
  ggplot2::ggplot(ggplot2::aes(x = frac_tokens, y = frac_words)) + 
  ggplot2::geom_point() 

```

Note that only about 5% of the morpheme tokens (out of about 194K) are needed to 
cover about 50% of the words (out of about 644K).

The plot above weights each word equally, no matter how frequently that word occurs.
To get a sense of word coverage by actual usage, we can weight each word by how often it occurs in some corpus. 
For example, weighting words by occurrence in the wikitext-103 corpus:

```{r count-weighted}
# passing a word frequency table in (columns: word, word_count) applies weights
# to the token counts. 
token_counts_weighted <- count_tokens(unnested_lookup, word_frequency_table)
utils::head(token_counts_weighted)
```

Now the top token is "the", just because of how common that word is.

Instead of counting how many words are covered by tokens up to some rank, we should count the *total weight* of words covered. 
For this, we join onto word_frequency_table again (giving words not found in the corpus a count of one).

```{r plot-weights}
words_with_max_token_ranks_weighted <- dplyr::left_join(
  unnested_lookup, 
  token_counts_weighted, 
  by = "token"
) %>% 
  dplyr::group_by(word) %>% 
  dplyr::summarize(max_rank = max(rank)) %>% 
  dplyr::filter(!is.na(max_rank))

weighted_tokens_and_words <- dplyr::left_join(
  words_with_max_token_ranks_weighted,
  word_frequency_table,
  by = "word"
) %>% 
  dplyr::mutate(
    word_count = ifelse(
      test = is.na(word_count), 
      yes = 1L, 
      no = word_count
    )
  )

words_vs_tokens_weighted <- weighted_tokens_and_words %>% 
  dplyr::group_by(max_rank) %>% 
  dplyr::summarize(n_words = sum(word_count)) %>% 
  dplyr::arrange(max_rank) %>% 
  dplyr::mutate(n_words = cumsum(n_words))

# plot!
words_vs_tokens_weighted %>% 
  dplyr::mutate(frac_words = n_words/max(n_words)) %>% 
  dplyr::mutate(frac_tokens = max_rank/max(max_rank)) %>% 
  ggplot2::ggplot(ggplot2::aes(x = frac_tokens, y = frac_words)) + 
  ggplot2::geom_point() 
```

The top 5% of tokens would cover over 98% of words by usage.

Now we make an actual vocabulary and lookup.

We construct a morphemepiece vocabulary, starting from a wordpiece vocabulary.
The constructed vocabulary contains the same "fancy" tokens as wordpiece (punctuation, etc.), but replaces most (non-short, non-proper-noun) words with their morpheme tokens. 
This process generally results in a core vocabulary with about 16K tokens. 

If our target vocabulary size is larger than this, additional tokens are added from the (ranked) list of tokens.

The lookup contains every word in the wikipedia word list which is covered by (processes into) the tokens in the vocabulary.

Here, let's make two vocabularies: one "small" (minimum to roughly cover the wordpiece vocabulary) and one "large" (comparable to the size of the wordpiece vocabulary).

```{r make-vocabs}
vandl_small <- make_vocab_and_lookup(
  full_lookup = original_lookup,
  full_vocabulary = wikimorphemes::wiktionary_word_list(),
  wordpiece_vocab = wordpiece.data::wordpiece_vocab(),
  target_vocab_size = 0, # no extra tokens
  word_frequency_table = word_frequency_table
)

vandl_large <- make_vocab_and_lookup(
  full_lookup = original_lookup,
  full_vocabulary = wikimorphemes::wiktionary_word_list(),
  wordpiece_vocab = wordpiece.data::wordpiece_vocab(),
  target_vocab_size = 30000L,
  word_frequency_table = word_frequency_table
)
```

The vocabulary and lookup aren't yet in our standardized forms.
First, we save them as standardized text files.

```{r make-lookups}
# TODO: Make this save and reload stuff unnecessary!
text_lookup_small <- .make_text_lookup(
  voc = vandl_small$vocab, 
  lu = vandl_small$lookup,
  word_freq_tab = word_frequency_table
)

writeLines(
  text_lookup_small, 
  file.path(morphemepiece_cache_dir(), "mp_lookup_small.txt")
)
# vocab is already just a character vector
writeLines(
  vandl_small$vocab, 
  file.path(morphemepiece_cache_dir(), "mp_vocab_small.txt")
)

# now do large
text_lookup_large <- .make_text_lookup(
  voc = vandl_large$vocab, 
  lu = vandl_large$lookup,
  word_freq_tab = word_frequency_table
)

readr::write_lines(
  text_lookup_large, 
  file.path(morphemepiece_cache_dir(), "mp_lookup_large.txt")
)
# vocab is already just a character vector
readr::write_lines(
  vandl_large$vocab, 
  file.path(morphemepiece_cache_dir(), "mp_vocab_large.txt")
)
vocab_raw <- readr::read_lines(
  file.path(morphemepiece_cache_dir(), "mp_vocab_large.txt")
)

# Read back from text files to process as standard {morphemepiece} files:
vocab <- load_or_retrieve_vocab(
  file.path(morphemepiece_cache_dir(), "mp_vocab_large.txt")
)
lookup <- load_or_retrieve_lookup(
  file.path(morphemepiece_cache_dir(), "mp_lookup_large.txt")
)

morphemepiece_tokenize("Surprisingly easy", vocab, lookup)
morphemepiece_tokenize("'Twas brillig, and the slithy toves", vocab, lookup)
```

## Check coverage

With these vocabularies, we can look at various measures of coverage.
For example, what (weighted) fraction of words from our corpus are covered by
the vocab/lookup?

```{r check-coverage-small}
corpus_coverage_small <- dplyr::left_join(
  word_frequency_table, 
  vandl_small$lookup, 
  by = "word"
) %>% 
  dplyr::mutate(covered_lookup = !is.na(tokenization)) %>% 
  # not every word in the vocab is in the lookup; check vocab too
  dplyr::mutate(covered_vocab = word %in% vandl_small$vocab) %>% 
  dplyr::mutate(covered = covered_lookup | covered_vocab) %>% 
  dplyr::mutate(covered_weighted = covered*word_count) 

corpus_coverage_small %>% 
  dplyr::summarize(sum(covered_weighted)/sum(word_count))
```

The "small" vocabulary covers 95.1% of words from the corpus, weighted by usage. 

```{r check-coverage-large}
corpus_coverage_large <- dplyr::left_join(
  word_frequency_table, 
  vandl_large$lookup, 
  by = "word"
) %>% 
  dplyr::mutate(covered_lookup = !is.na(tokenization)) %>% 
  # not every word in the vocab is in the lookup; check vocab too
  dplyr::mutate(covered_vocab = word %in% vandl_large$vocab) %>% 
  dplyr::mutate(covered = covered_lookup | covered_vocab) %>% 
  dplyr::mutate(covered_weighted = covered*word_count) 

corpus_coverage_large %>% 
  dplyr::summarize(sum(covered_weighted)/sum(word_count))
```

The "large" vocabulary covers 96.6% of words, 1.5% more words (weighted) than the "small" vocabulary. 

It is useful to look at the most common words that are *not* covered by the vocabularies:

```{r show-coverage-large}
#large
uncovered <- corpus_coverage_large %>% 
  dplyr::filter(!.data$covered) %>% 
  dplyr::arrange(dplyr::desc(.data$word_count)) %>% 
  dplyr::select(.data$word, .data$word_count) %>% 
  head(100) %>% 
  dplyr::mutate(
    tokenization = morphemepiece_tokenize(
      .data$word, 
      vocab = vocab, 
      lookup = lookup
    )
  ) %>% 
  dplyr::rowwise() %>% 
  dplyr::mutate(
    tokenization = paste(names(.data$tokenization), collapse = " ")
  ) %>% 
  dplyr::ungroup()

head(uncovered, 10)
```

Categories of uncovered words include some proper nouns and partial contractions (like "didn").

The word "los" in this list raises an interesting question for future consideration: should a small set of very common words in languages besides English be included? After all, a significant number of tokens in both wordpiece and morphemepiece are currently used for Asian language characters.

## Other measures of coverage quality

Some other checks we will try include:

- For the top 1000 common words (e.g. as estimated [here](https://xkcd.com/simplewriter/words.js)), manually evaluate the quality of the breakdown.
- For N random covered words *without* a breakdown, manually evaluate whether the word should have a breakdown.
  
(These checks have not been re-ran yet with the updated data.)

```{r xkcd, eval = FALSE}
# just for fun :-D
xkcd_words_url <- "https://xkcd.com/simplewriter/words.js"
raw_words <- readr::read_lines(xkcd_words_url)

raw_words <- raw_words[grepl("WORDS", raw_words)]
raw_words <- stringr::str_split(raw_words, '"')[[1]]
raw_words <- raw_words[grepl("\\|", raw_words)]
words <- dplyr::tibble(top_words = stringr::str_split(raw_words, "\\|")[[1]])
# I feel lied to. There are more than 3k words in this list.
words <- words %>% 
  dplyr::mutate(
    tokenized = morphemepiece_tokenize(.data$top_words, vocab, lookup)
  ) %>% 
  dplyr::rowwise() %>% 
  dplyr::mutate(
    tokenized = paste(names(.data$tokenized), collapse = " ")
  ) %>% 
  dplyr::ungroup()

# ss_url <- "url of mp_scratch google sheet"

# already authorized
# googlesheets4::write_sheet(words, ss_url)
# manual check, add column "is_ok"

# checked_words <- googlesheets4::read_sheet(ss_url, sheet = "check common words")

# if breakdown is ok, value is "y"

# mean(checked_words$is_ok == "y")

# [1] 0.9711062
# many of the exceptions can/should be fixed in wiktionary
# These have not been checked in a while.
```

Fewer than 3% of these words (about 100 total) were judged to be problematic.
For these common words, it makes sense for us to address the issues directly in Wiktionary where possible.

For the next check, randomly pick some words of at least length 4 without a 
breakdown:

```{r more-checks, eval = FALSE}
all_words <- unique(unnested_lookup$word)
unbroken_vocab_words <- intersect(names(vocab), all_words)
unbroken_vocab_words <- unbroken_vocab_words[nchar(unbroken_vocab_words) > 3]

# sample a few hundred
unbroken_sample <- dplyr::tibble(unbroken_word = sample(unbroken_vocab_words, 
                                                        size = 300))
# send to google sheet for manual check

# googlesheets4::write_sheet(unbroken_sample, ss_url, sheet = "unbroken_check")
# manual check, add column "is_ok"
# checked_unbroken_words <- googlesheets4::read_sheet(
#   ss_url, 
#   sheet = "check unbroken words"
# )

# if breakdown is ok, value is "y"
# table(checked_unbroken_words$is_ok)
 #  ?   n   y 
 # 24  32 244 
# many of the exceptions can/should be fixed in wiktionary
```

About 10% of this sample was judged to be wrong (definitely should have a breakdown), and about 10% was judged to be questionable (maybe should have a breakdown).
